Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-09-30 Manuscript/manuscriptPS.tex
Update manuscript

\subsection{Counterion binding to lipid bilayers containing PS lipids}
Membranes containing PS lipids are always accomppanied with counterions, which
modulate electrostatic interactions between lipids and other biomolecules. 
Counterions are also suggested screen the repulsion between charged lipid headgroups 
in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
The counterion density profiles along membrane normal 
show significant differences between simulation models (Fig. \ref{NAdensPOPS}).
The strongest counterion binding, i.e., the lowest concentrations in bulk water,
are observed in MacRog, Berger and Lipid17/JC simulations.
CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
density, while a single maxima is observed in the other models. 
\todo{More detailed discussion may be possible after comparing monovalent ion binding
to bilayers between CHARMM simulations and experiments.
Also, section S6 should be finished.}
Area per lipid is in agreement with experiments \cite{pan14} only
in the Gromos-CKP models, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
The difference cannot be explained by the electrostatic screening of the headgroup repulsion due to 
counterion binding because CHARMM36, CHARMM36ua and Slipid models give
smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
%The proposed trend is observed for Lipid17 model,
%for which the Joung-Cheatham ions give higher affinity and smaller area
%per molecule. However, the trend is not observed when comparing accross different
%force fields. For example, CHARMM36ua and MacRog give similar area per molecule
%but binding affinity is significantly higher in MacRog. 
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}

To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
as a function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
Experimental order parameter data for POPC headgroup in the mixture is available as a function
of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
as a function of NaCl. Lithium interacts more strongly with PS headgroups than other monovalent 
ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
This is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
but only modest changes as a function of KCl concentration, indicating singificant 
Li$^+$ binding but only weak Na$^+$ binding to the mixture when interpreted using the
electrometer concept~\cite{akutsu81,altenbach84,seelig87}. In simulations with the
Berger model, the headgroup order parameter response of POPC
to the added NaCl is similar to the experiments of LiCl,
indicating overestimated binding affinity of sodium, in line with the results for PC bilayers \cite{catte16}.
Indeed, the sodium density profile shows a significant binding peak in the 
Berger model (Fig. \ref{CIdensPSOCmixt}). Potassium binding in the MacRog simulation
is significantly weaker  (Fig. \ref{CIdensPSOCmixt}) and the headgroup order parameter 
changes are also in better agreement with simulations (Fig. \ref{PSresponseTONaCl}).
\todo{Discussion about Lipid17 to be written when we have the density profiles.}
All the tested models overestimate the changes of POPS headgroup order parameters as
a function of monovalent ions (Fig. \ref{PSresponseTONaCl}), suggesting that
model development is necessary to interpret the PS headgroup-ion interactions 
from MD simulations.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture at 298~K
    (except Berger simulations are (4:1) mixture at 310~K).
    The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Counterion density distributions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
  \todo{Lipid 17 is to be added when we have the density profiles.
  There is something wrong with the current plots in git: https://github.com/NMRLipids/MATCH/pull/40}
\end{figure}

\subsection{Headgroup structure in PS and PC mixtures}
Dilution of PS lipid bilayers with PC lipids reduces the propensity
of PS headgroup-multivalent ion complexes and is proposed to make PS
headgroups less rigid~\cite{browning80,buldt81,roux90,roux91}.
Therefore, the intermolecular interactions at the headgroup region seems
to be important for the physical properties of mixed lipid bilayers.
These interactions can be indirectly monitored by measuring the
headgroup order parameters from PS:PC mixtures with different molar
ratios. The headgroup order parameters of POPC increase in such experiments 
with increasing amount of POPS (Fig.~\ref{HGorderparametersPCvsPS})~\cite{scherer87}. 
This behaviour is generally observed when negatively charged lipids or surfactants 
are mixed with PC lipids~\cite{scherer87,scherer89} and can be understood by the 
tilting of lipid headgroup more parallel to the membrane plane according to
the electrometer concept \cite{seelig87}. The headgroup order parameters of PS lipids
shift closer to zero when bilayer is diluted with PC lipids in 
experiments (Fig.~\ref{HGorderparametersPCvsPS})~\cite{browning80,scherer87,roux90},
which is intepreted to indicate reduced rigidity~\cite{browning80,buldt81}.

The increase of POPC headgroup order parameters with the increasing
amount of negatively charged POPS lipid is reproduced in
MacRog simulations with potassium counterions, but not in 
Berger simulations with sodium or in CHARMM36 simulations
with potassium or sodium conterions (Fig. \ref{HGorderparametersPCvsPS}). 
The observations can be explained using the electrometer concept.
The Berger simulation exhibits very strong sodium binding (Fig. \ref{CIdensPSOCmixt}), 
which surpasses the effect of negatively charged lipids as also the amount of
counterions increase with increasing amount of PS.
In CHARMM36 simulations, the counterion binding neutralizes the effect
of PS and the headgroup order parameters are not changed with increasing amount
of PS. Finally, the weak binding of potassium in the MacRog simulations
enables the increase of order parameters with the increasing amount of
negatively charged PS lipids (Figs. \ref{HGorderparametersPCvsPS} and \ref{CIdensPSOCmixt}).

\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of PC (left panel) and PS (right panel) headgroup order parameters
    from POPC:POPS mixtures with increasing amount of POPS.
    Experimental results of POPC are taken from Ref. \citenum{scherer87}
    (signs are determined as discussed in \cite{botan15,ollila16}).
    Experimental values for POPS in pure bilayer and in mixture are  
    measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
    Since the experimental data of POPS in pure and diluted mixture come from
    different experimental sets (13C NMR in this work and 2H NMR from Ref. \citenum{roux90}),
    the experimental change of the order parameter is less accurate than in typical measurements
    where same technique is used in all conditions, see discussion about qualitative and quantitative 
    accuracy in Ref. \citenum{ollila16}.
    For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
    For $\beta$-carbon of POPS (right panel, top) the zero point of y-axis is
    set to the value from POPC:POPS (5:1) mixture.
    For $\alpha$-carbon of POPS (right panel, bottom) the y-axis is tranferred
    with the same value for both order parameters such that the lower order
    parameter value from POPC:POPS (5:1) mixture is at zero
    to correctly illustrate the significant forking.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{Simulation of pure POPC at 298K with Lipid14 would be useful for this plot (only at 303 K is available from NMRlipids I)} \\
  \todo{MacRog simulations of pure POPS with potassium counterions only would be useful for this and other plots.} \\
  \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
\end{figure*}

>>>>

\subsection{Counterion binding and interactions between PC and PS headgroups}
Membranes containing PS lipids are always accomppanied with counterions which
modulate electrostatic interactions between lipids and other biomolecules. 
Counterions are also suggested to screen the repulsion between charged lipid headgroups 
in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
The counterion density profiles along membrane normal 
show significant differences between simulation models in both binding affinity
and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
The strongest sodium counterion binding, i.e., the lowest concentrations in bulk water,
is observed in MacRog, Berger and Lipid17/JC simulations.
Slipids, Berger, MacRog, and Lipid17 exhibit a clear single maxima in the
counterion density, but slightly at different distances from the bilayer
center. Other models exhibit more complicated distribution, often including
two local maxima in the density profile.
%CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
%density, while a single maxima is observed in the other models.

Area per lipid is in agreement with experiments \cite{pan14} only
in the MacRog simulation with potassium counterions and in Gromos-CKP models,
while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
The difference cannot be explained only by the electrostatic screening of the headgroup repulsion due to 
counterion binding because CHARMM36, CHARMM36ua and Slipid models give
smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
On the other hand, increasing counterion binding affinity reduces the area per molecule
when comparing between sodium and potassium in MacRog simulations or between
ff99 and jc ions in Lipid 17 simulations.
This suggests that the low area per molecule in PS lipid bilayers cannot be
explained only by counterion binding but also the hydrogen bonding network has to
be taken into account, as also previously suggested based on scattering experiments \cite{petrache04}.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}

Previously used method to evaluate counterion binding affinity to lipid bilayers
using the changes of hedgroup order parameters as a function
additional ions \cite{catte16} is difficult to apply for counterions because
reference state without ions in solution does not exist, see section \ref{mixtureTOadditionalCIs}
in supplementary information. However, the counterion binding affinity is
reflected to the increase of POPC headgroup order parameters due to the
increasing amount of negatively charged POPS lipids in the bilayer.
%The headgroup order parameters of POPC increase
%with increasing amount of negatively charged POPS
The experimentally measured order parameter increase~\cite{scherer87}
is reproduced in simulations with MacRog and potassium counterions, but not in 
other simulations (Fig. \ref{HGorderparametersPCvsPS}). 
The increase of POPC headgroup order parameters with increasing amount of POPS 
can be explained using the electrometer concept \cite{seelig87,scherer89}:
The PC headgroup tilts more parallel to the membrane plane with increasing
amount of negative charge in the membrane. In Berger simulations,
the strong sodium binding surpasses the effect of negatively charged PS
lipids and the order parameters actually decrease because also the amount of
positive counterions increase with increasing amount of PS.
In CHARMM36 simulations, the counterion binding neutralizes the effect
of PS and the headgroup order parameters are not changed with increasing amount
of PS. Therefore, we conclude that the weak binding of potassium in the MacRog
simulations (Fig. \ref{NAdensPOPS}) leads to the correct surface charge
density and the response of POPC headgroup to the increasing amount of negatively
charges PS lipids in the lipid mixtures. 
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of PC (left panel) and PS (right panel) headgroup order parameters
    from POPC:POPS mixtures with increasing amount of POPS.
    Experimental results of POPC are taken from Ref. \citenum{scherer87}
    (signs are determined as discussed in \cite{botan15,ollila16}).
    Experimental values for POPS in pure bilayer and in mixture are  
    measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
    Since the experimental data of POPS in pure and diluted mixture come from
    different experimental sets (13C NMR in this work and 2H NMR from Ref. \citenum{roux90}),
    the experimental change of the order parameter is less accurate than in typical measurements
    where same technique is used in all conditions, see discussion about qualitative and quantitative 
    accuracy in Ref. \citenum{ollila16}.
    For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
    For $\beta$-carbon of POPS (right panel, top) the zero point of y-axis is
    set to the value from POPC:POPS (5:1) mixture.
    For $\alpha$-carbon of POPS (right panel, bottom) the y-axis is tranferred
    with the same value for both order parameters such that the lower order
    parameter value from POPC:POPS (5:1) mixture is at zero
    to correctly illustrate the significant forking.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{Simulation of pure POPC at 298K with Lipid14 would be useful for this plot (only at 303 K is available from NMRlipids I)} \\
  \todo{MacRog simulations of pure POPS with potassium counterions only would be useful for this and other plots.} \\
  \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
\end{figure*}

The headgroup order parameters of PS lipids
shift closer to zero when bilayer is diluted with PC lipids in 
experiments (Fig.~\ref{HGorderparametersPCvsPS})~\cite{browning80,scherer87,roux90},
which is intepreted to indicate reduced rigidity~\cite{browning80,buldt81}.

Dilution of PS lipid bilayers with PC lipids reduces the propensity
of PS headgroup-multivalent ion complexes and is proposed to make PS
headgroups less rigid~\cite{browning80,buldt81,roux90,roux91}.
Therefore, the intermolecular interactions at the headgroup region seems
to be important for the physical properties of mixed lipid bilayers.
These interactions can be indirectly monitored by measuring the
headgroup order parameters from PS:PC mixtures with different molar
ratios.

### 2018-09-30 Manuscript/manuscriptPSsuppl.tex
Update manuscript

\section{Sodium binding to DMPC:DOPS mixture}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PSresponseTONaCl.eps}
  \caption{\label{PSresponseTONaClDMPC}
    Order parameters of PS headgroup as a function of added NaCl measured from DMPC:DMPS (3:1) mixture \cite{roux86}.
  }
\end{figure}
The experimental results show essentially no changes in the order parameters as a function of
added NaCl, while significant changes are observed in simulations. However,
the minimum buffer concentration of NaCl in the experimental was 100mM \cite{roux86}.
Therefore, we cannot exclude the possibility that the NaCl induced changes were already
saturated with 100mM NaCl concentration, which was the case for CaCl$_2$ in Fig. \ref{changesWITHCaClPS}.
\pagebreak

>>>>

\section{Headgroup response to the additional counterions in POPC:POPS (5:1) mixtures}\label{mixtureTOadditionalCIs}
To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
as a function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
Experimental order parameter data for POPC headgroup in the mixture is available as a function
of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
as a function of NaCl. Lithium interacts more strongly with PS headgroups than other monovalent 
ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
This is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
but only modest changes as a function of KCl concentration, indicating singificant 
Li$^+$ binding but only weak Na$^+$ binding to the mixture when interpreted using the
electrometer concept~\cite{akutsu81,altenbach84,seelig87}. In simulations with the
Berger model, the headgroup order parameter response of POPC
to the added NaCl is similar to the experiments of LiCl,
indicating overestimated binding affinity of sodium, in line with the results for PC bilayers \cite{catte16}.
Indeed, the sodium density profile shows a significant binding peak in the 
Berger model (Fig. \ref{CIdensPSOCmixt}). Potassium binding in the MacRog simulation
is significantly weaker  (Fig. \ref{CIdensPSOCmixt}) and the headgroup order parameter 
changes are also in better agreement with simulations (Fig. \ref{PSresponseTONaCl}).
\todo{Discussion about Lipid17 to be written when we have the density profiles.}
All the tested models overestimate the changes of POPS headgroup order parameters as
a function of monovalent ions (Fig. \ref{PSresponseTONaCl}), suggesting that
model development is necessary to interpret the PS headgroup-ion interactions 
from MD simulations.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture at 298~K
    (except Berger simulations are (4:1) mixture at 310~K).
    The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Counterion density distributions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
  \todo{Lipid 17 is to be added when we have the density profiles.
  There is something wrong with the current plots in git: https://github.com/NMRLipids/MATCH/pull/40}
\end{figure}

%%
%% I think that this section is not very useful, I think that we should remove it.
%%
%\section{Sodium binding to DMPC:DOPS mixture}
%
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/PSresponseTONaCl.eps}
%  \caption{\label{PSresponseTONaClDMPC}
%    Order parameters of PS headgroup as a function of added NaCl measured from DMPC:DMPS (3:1) mixture \cite{roux86}.
%  }
%\end{figure}
%The experimental results show essentially no changes in the order parameters as a function of
%added NaCl, while significant changes are observed in simulations. However,
%the minimum buffer concentration of NaCl in the experimental was 100mM \cite{roux86}.
%Therefore, we cannot exclude the possibility that the NaCl induced changes were already
%saturated with 100mM NaCl concentration, which was the case for CaCl$_2$ in Fig. \ref{changesWITHCaClPS}.
\pagebreak
